Cheminformatics: The Computer Science of Chemical Discovery

### 2011-11-10 rebuttal.tex

\comment{page 1, column 1, line 24: the authors need to provide a
  reference for the assertion that cheminformatics is an older field,
  particularly given the comments later in this
  manuscript.}
\response{We have removed the assertion that cheminformatics is older
  than bioinformatics. Primarily, because there is no obvious
  reference that claims this. But alse because, depending on what one
  considers bioinformatics and cheminformatics, one field can be
  identified as being older or younger than the other. We feel that
  claims on age do not strengthen the paper and have thus removed it}

\comment{p1, c1, l29-37: I would dispute that until recently the
  cheminformatics techniques have been closely guarded
  secrets. J. Chem. Doc. was founded in 1961 and other techniques even
  older, such as the Wiener index (1947), Wiswesser Line Notation
  (1949), etc. Not to mention the actual foundations of what we call
  cheminformatics back to the atomistic theory of the 19th century.}
\response{We have the updated the text on page 1 to attenuate this
  statement, stressing on the fact that much of the data in
  cheminformatics is proprietary, whereas only some of the techniques (such
  as SMILES canonicalization) are. Regarding the statement that atomistic
  theory laid the foundations of cheminformatics - this is certainly
  true in a very broad sense. For that matter the atomistic theory
  laid the ground for computational chemistry in general. In the
  supplemental history document, we make a distinction between
  computational chemistry and cheminformatics}

\comment{p1, c1, l32: I think the phrase miracle molecule could be
  more usefully replaced with new drugs or new small molecule
  therapeutics.}
\response{The text has been updated to use the term ``therapeutic
  molecule''}

\comment{p2, c2, l57: it might be worthwhile here clarifying in the text
precisely the properties that require satisfaction to deliver a small
molecule therapeutic, not ?promising?. A drug must be safe and
efficacious ultimately; perhaps this should be mentioned first
followed by the typical pitfalls and how they are assessed?}
\response{The text has been updated to provide some examples of
  properties that would characterize a therapeutically useful small molecule}

### 2011-11-10 paper.tex

\section{The Role of Open Source \& Open Data}
\label{sec:role-open-source}

A key aspect of cheminformatics is that much of its development
requires access to collections of molecular structures (and, usually,
associated bioactivity data). As has been noted, it is only recently
that large structure and bioactivity data collections have become
available. As a result, this enables much more robust validation of
cheminformatics methodologies as well as large scale
\emph{benchmarking of algorithms}. The latter is especially relevant
for the plethora of data mining techniques that are employed in
cheminformatics. At this point, the nature of Open Data focuses
primarily on structure and activity data types. On the other hand,
there is a distinct lack of textual data (such as journal articles)
that are ``Open'' in nature. While PubMed abstracts serve as a proxy
for journal articles, text mining methods in cheminformatics are
hindered by not being able to mine the full text of many scientific
publications.  Interestingly, patent information is publicly
accessible (cf. Google Patents) and there have been efforts to mine
these for chemical information using Open tools.

Obviously freely accessible data is important for research; but
equally important are the tools to work with chemical data---more
specifically, chemical structures. Chemical toolkits are the workhorses
of cheminformatics research and a variety of toolkits are available,
both commercial (which, in many cases, provide free academic
licenses) and Open Source. In this article we do not debate the
arguments for and against Open Source; rather, we focus on Open Source
tools, since this ensures that a reader will be able to explore
cheminformatics problems with a minimum of encumbrances.

\subsection{Open Source Chemical Toolkits}
\label{sec:toolkits}

The extent of open-source software in cheminformatics has somewhat
lagged behind related fields such as bioinformatics, partly
due to the existence of a lucrative pharmaceutical market for
proprietary software. However, in recent years there has been a
rise in the amount of open-source cheminformatics software. One
contribution to this has been the Blue Obelisk group
(\url{http://blueobelisk.org}) of chemists, who have worked together to
create interoperable open-source chemistry software (among other
activities). Now there are a number of open-source toolkits available
including the Chemistry Development
Kit (CDK)~\cite{steinbeck2003}, Open Babel, RDKit and Indigo. These toolkits are written in
Java or C++ (though toolkits written in the latter invariably have
bindings to a variety of languages). Such toolkits are used, for
example, to read/write chemical file formats, manipulate chemical
structures, measure similarity of molecules, search for substructures,
and generate 2D diagrams. The underlying algorithms used include graph
algorithms (e.g. maximal common substructure, canonical labelling of
graphs), geometrical methods (e.g. Kabsch alignment) and vector
manipulation (e.g. converting coordinates in various systems, 3D
structure generation). As well as being of direct use to practising
cheminformaticians in their day-to-day work, many applications have
been developed that rely on these toolkits to handle chemical data;
for example, the molecular viewer Avogadro
(\url{http://avogadro.openmolecules.net}) uses Open Babel, while the
molecular workbench Bioclipse \cite{Bioclipse2} uses the CDK.

The various toolkits have many features in common and at the same time
have certain distinguishing features. For example, the CDK implements
a large collection of molecular descriptors. On the other hand,
Open Babel has robust support for 3D structure generation and
optimization and supports the interconversion of a huge number of
chemical file formats.  Table~\ref{tab:features} (in the Appendix) compares features offered by the open-source toolkits currently available.

While algorithmic challenges are certainly faced by toolkits, software
engineering challenges probably affect the use and development of the
toolkits on a daily basis. Examples include verifying code coverage,
thread safety, documentation quality and so on. Some projects such as
the CDK employ a variety of tools to address some of these issues.
Examples include the use of PMD to analyze code quality and JUnit to
perform unit testing. At the level of implementation, there are still
inefficiencies in some toolkits. For example the CDK employs a JJTree
parser to implement the SMARTS specification, but employs a hand built
parser to handle SMILES. The latter is essentially a complex tangle of
\emph{if \ldots then} statements and is not easily maintainable.

It is also important to note that certain challenges faced by nearly
all cheminformatics toolkits stem from the graph representation of
chemistry. This representation, while succinct and amenable to
computation, is only an approximation of reality---and edge cases
abound. Existing areas of interest are enumeration of colored graphs,
taking into account symmetry, and symmetry detection itself, which not
merely derives from the chemical graph, but typically also from the 3D
geometry of the chemical. However, more challenging is that chemical
representation of a molecule really needs to capture non-static
graphs, in order to take into account delocalisation and the tautomeric
phenomena molecules can undergo in practice. It is important to
realize that although chemical representations have increased in
sophistication over the years, many chiral molecules cannot be
satisfactorily handled by current representation systems.

Given that there is ample scope for software engineering (algorithm
implementation, code quality and analysis, build systems etc), both
the CDK and Open Babel are open to new contributions. Both are driven
via public, active mailing lists and public issue tracking systems.
Development questions are actively and openly discussed to arrive at a
consensus and contributors are encouraged to get involved with any of
the following areas: development of new features, improvement of
existing features, performance optimization, fixing bugs,
documentation, website development, reporting bugs, and helping users
on the mailing list. Anyone interested should email the respective
developers' mailing list (either \texttt{cdk-devel@lists.sf.net} or
\texttt{openbabel-devel@lists.sf.net}) with their interests and
background, and a mentor will be assigned to help them join the
project.
%

>>>>

\section{Support for the Development of Novel Algorithms}
\label{sec:development-support}

For researchers interested in the development of new algorithms in the
field, a software library that handles chemical structures and related
data is essential. A wide variety of such chemical toolkits are
available, both commercial (which may offer free academic
licenses) and Open Source.

Due to the importance of cheminformatics to the pharmaceutical
industry in particular, numerous commercial vendors provide libraries,
applications and database cartridges to handle a wide variety of
needs including Accelrys, BioSolveIT, ChemAxon, Chemical Computing Group,
Daylight, Open Eye, Schrodinger, Tripos, and Xemistry. More recently
there has been a growth in the development of Open
Source cheminformatics software, and this presents opportunities for
the rapid development and implementation of novel algorithms which can
build on the existing Open Source ecosystem. In this article we do not debate the
arguments for and against Open Source; we focus on Open Source
tools simply because this ensures that a reader will be able to explore
cheminformatics problems with a minimum of encumbrances.

The extent of open-source software in cheminformatics has somewhat
lagged behind related fields such as bioinformatics, partly
due to the existence of a lucrative pharmaceutical market for
proprietary software. However, in recent years there has been a
rise in the amount of open-source cheminformatics software. One
contribution to this has been the Blue Obelisk
group~\cite{BlueObelisk2011}
(\url{http://blueobelisk.org}) of chemists, who have worked together to
create interoperable open-source chemistry software (among other
activities). Now there are a number of Open-Source chemical toolkits available
including the Chemistry Development
Kit (CDK)~\cite{steinbeck2003}, Open Babel~\cite{openbabel2011}, RDKit and Indigo. These toolkits are written in
Java or C++ (though toolkits written in the latter invariably have
bindings to a variety of languages). Such toolkits are used, for
example, to read/write chemical file formats, manipulate chemical
structures, measure similarity of molecules, search for substructures,
and generate 2D diagrams. The underlying algorithms used include graph
algorithms (e.g. maximal common substructure, canonical labelling of
graphs), geometrical methods (e.g. Kabsch alignment) and vector
manipulation (e.g. converting coordinates in various systems, 3D
structure generation). As well as being of direct use to practising
cheminformaticians in their day-to-day work, many applications have
been developed that rely on these toolkits to handle chemical data;
for example, the molecular viewer Avogadro
(\url{http://avogadro.openmolecules.net}) uses Open Babel, while the
molecular workbench Bioclipse~\cite{Bioclipse2} uses the CDK.

The various toolkits have many features in common and at the same time
have certain distinguishing features. For example, the CDK implements
a large collection of molecular descriptors. On the other hand,
Open Babel has robust support for 3D structure generation and
optimization and supports the interconversion of a huge number of
chemical file formats. Table~\ref{tab:features} (in the Appendix) compares features offered by the open-source toolkits currently available.
Given that there is ample scope for software engineering (algorithm
implementation, code quality and analysis, build systems etc), both
the CDK and Open Babel are open to new contributions. Both are driven
via public, active mailing lists and public issue tracking systems.

It is important to note that certain challenges faced by nearly
all cheminformatics toolkits stem from the graph representation of
chemistry. This representation, while succinct and amenable to
computation, is only an approximation of reality---and edge cases
abound. Existing areas of interest are enumeration of colored graphs,
taking into account symmetry, and symmetry detection itself, which not
merely derives from the chemical graph, but typically also from the 3D
geometry of the chemical. It is important to
realize that although chemical representations have increased in
sophistication over the years, many chiral molecules cannot be
satisfactorily handled by current representation systems. Perhaps a more challenging aspect is that
the chemical representation of a molecule really needs to capture non-static
graphs, in order to take into account delocalisation and the tautomeric
phenomena molecules can undergo in practice.

Of equal importance to the development of new algorithms is freely
available data on which to train and test new methods. As has been noted, recently
large structure and bioactivity data collections have become
available. As a result, this enables much more robust validation of
cheminformatics methodologies as well as large scale \emph{benchmarking of
algorithms}. The latter is especially relevant for the plethora of
data mining techniques that are employed in cheminformatics. At this
point, the nature of Open Data focuses primarily on structure and
activity data types. On the other hand, there is a distinct lack of
textual data (such as journal articles) that are ``Open'' in nature. While
PubMed abstracts serve as a proxy for journal articles, text mining
methods in cheminformatics are hindered by not being able to mine the full text of many scientific publications.
Interestingly, patent information is publicly accessible (cf. Google
Patents) and there have been efforts to mine these for chemical
information using Open tools.
